---
title: "Example Analysis - UFC Database"
author: "Frankie Noguera"
format: html
editor: visual
---

# Example Analysis Notebook

### Welcome new Data Scientist, I am glad you took the time to look at my database. Here I will show you guide you through how to use my database. As well, I will show you how to answer some potential questions so you can be inspired and then give it a try! Let's get into it!

## STOP! BEFORE MOVING ON, IMPORT DATASETS & PACKAGES:

```{r}
# Install packages
install.packages("tidyverse") # include " "
install.packages()
```

```{r}
# Library packages in
library(tidyverse) #Do not incoude " "
```

```{r}
# Importing the datasets. Remember that the CSV's should be in the same directory or read it in the directory where you save your CSV's
read_csv("ufc_fighters_stats.csv") -> ufc_fighters_stats #assign name you want for table
read_csv("ufc_country_data.csv") -> ufc_country_data #assign name you want for table
read_csv("ufc_events_stats.csv") -> ufc_events_stats #assign name you want for table
```

## ONE MORE STEP! LOL MY BAD! BUT ALWAYS OBSERVE YOUR DATASET AND TRY TO UNDERSTAND IT.

```{r}
ufc_fighters_stats
```

```{r}
ufc_country_data
```

```{r}
ufc_events_stats
```

### We will also be using ggplot(). What is ggplot?

-   ggplot is an R package for creating visually appealing, customizable data visualizations. It uses a layered approach, where you define the data, map variables to visual properties (like x/y axes), and add layers (like points, lines, or bars). It’s part of the **tidyverse** and great for making charts like scatterplots, bar plots, and more.

### How to use ggplot?

#### first call on ggplot() then these are the parameters/arguments:

#### data  - The dataset you want to observe  aes() -  Aesthetics. Inside input comparison of x and y data  geom_point() - What type of plot. In this case a point plot  geom_smooth() - Here we visualize the line of best fit to see the trend  labs()  - How we the user want to label the x, y and the title

## 1. Question: Does a fighter's age impact their win percentage?

Answer: Yes, the older the more L's. We see a clear negative trend with this scatter plot (sorry Mr. Tyson)

```{r}
ggplot(data = ufc_fighters_stats, aes(x = age, y = win_percent)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Win Percentage vs. Age", x = "Age", y = "Win Percentage")
```

## 2. Question: Which weight division has fighters with the highest number of wins?

answer: looks like it's Heavyweight Division

bonus: find out how many wins

```{r}
#this is a box plot and fill is what color we want the inside of our boxes to be. Since I put weight divsion it would make a different color per weight division
ggplot(data = ufc_fighters_stats, aes(x =weight_division, y=wins, fill = weight_division)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = "Wins by Weight Division", x = "Weight Division", y ="Number of Wins")
```

## 3. Which country has the most amount of wins?

answer: United States

```{r}
ufc_fighters_stats |> #loading in the data
  group_by(country) |> #grouping data by the country
  summarize(total_wins = sum(wins, na.rm = TRUE)) |> #here we create a new column "total_wins"
  arrange(desc(total_wins)) -> country_wins #here we arrange it for highest wins on top
```

```{r}
ggplot(country_wins, aes(x = country, y = total_wins)) + 
  geom_col(fill = "blue") + #fill is what color we want to fill the columns with
  coord_flip() + # swaps x and y axes for a horizontalchart
  ggtitle("Total Wins by Country") + #the title will always be at the bottom
  xlab("Country") + # remember since this is flipped this will be to the left
  ylab("Total Wins") # since this is flipped it will be on the top
```

## 4. Which country has the highest total attendance across UFC events?

```{r}
# Here I am joining 2 tables by the country (ufc event and ufc conutry tables)
ufc_events_stats |>
  left_join(ufc_country_data, by = "country") -> event_country_data #saving result as this
```

```{r}
event_country_data |>
  group_by(country) |>
  summarize(total_attendance = sum(attendance, na.rm = TRUE)) |> # creating a new column based on only the sum of the attendance by country
  arrange(desc(total_attendance)) -> country_attendance #highest attendance/country first
```

```{r}
ggplot(country_attendance, aes(x = country, y = total_attendance)) +
  geom_point(size = 4, color = "darkblue") + #here we made a point plot to visualize
  coord_flip() +
  ggtitle("Total Attendance by Country") +
  xlab("Country") +
  ylab("Attendance")

```

## 5. Which country has the tallest average fighters in terms of height?

answer: Austria

```{r}
# joining both fighters and country datasets
ufc_fighters_stats |>
  left_join(ufc_country_data, by = "country") -> fighters_country_join
```

```{r}
fighters_country_join |>
  group_by(country) |> #observing this by country
  summarize(average_height = mean(height_ft, na.rm = TRUE)) |> #here we are creating a new column that finds the average height by country
  arrange(desc(average_height)) -> average_height_by_country #from tallest to shortest then saving it in average_height_by_country
```

```{r}
average_height_by_country |>
  slice(1) #isolating row 1
```

```{r}
ggplot(average_height_by_country, aes(x = country, y = average_height)) +
  geom_bar(stat = "identity", fill = "red", width = 0.7) + # here we made a bar chart and made it's width slimmer to be able to see most country names
  coord_flip()+
  ggtitle("Average Fighter Height by Country") +
  xlab("Country") +
  ylab("Average Height (ft)")
```
